### 2015-12-11 introduction.tex
Expand introduction

A summary of algorithms for OLO is presented in Table~\ref{table:results}.
While the scale-free property has been looked at in the expert setting, in the
general OLO setting this issue has been largely ignored.  In particular, the
\textsc{AdaHedge}~\cite{de-Rooij-van-Erven-Grunwald-Koolen-2014} algorithm, for
prediction with expert advice, is specifically designed to be scale-free.  A
notable exception in the OLO literature is the discussion of the ``off-by-one''
issue in~\cite{McMahan-2014}, where it is explained that even the popular
\textsc{AdaGrad} algorithm~\cite{Duchi-Hazan-Singer-2011} is not completely
adaptive; see also our discussion in Section~\ref{section:solo-ftrl}. In
particular, existing scale-free algorithms cover only some norms/regularizers
and \emph{only} bounded decision sets. The case of \textbf{unbounded decision
sets}, practically the most interesting one for machine learning applications,
remains completely unsolved.

Rather than trying to design strategies for a particular form of loss vectors
and/or decision sets, in this paper we explicitly focus on the scale-free
property. The regret of scale-free algorithms is proportional to the scale of the
losses, ensuring optimal linear dependency on the maximum norm of the loss
vectors.

We design and analyze three scale-free algorithms. First two algorithms, which
we call \textsc{AdaFTRL} and \textsc{SOLO FTRL}, are instances of
\textsc{Follow The Regularized Leader} (\textsc{FTRL}) with adaptive regularizers.
\textsc{AdaFTRL} is a generalization of
\textsc{AdaHedge}~\cite{de-Rooij-van-Erven-Grunwald-Koolen-2014} to arbitrary
strongly convex regularizers.  \textsc{SOLO FTRL} can be viewed as the
``correct'' scale-free version of \textsc{AdaGrad
FTRL}~\cite{Duchi-Hazan-Singer-2011} generalized to arbitrary strongly convex
regularizers.  The third algorithm, which we call \textsc{Scale-Free Mirror
Descent}, is a variant of \textsc{Mirror Descent}. It is a generalization of
\textsc{AdaGrad MD}~\cite{Duchi-Hazan-Singer-2011} to arbitrary strongly convex
regularizers, similar to \textsc{Optimistic MD}~\cite{Rakhlin-Sridharan-2013}.
The three algorithms are presented and analyzed in
Sections~\ref{section:ada-ftrl}, \ref{section:solo-ftrl} and
\ref{section:mirror-descent} respectively.

For all three algorithms, we prove that for bounded decision sets the regret
after $T$ rounds is at most $O \left(\sqrt{\sum_{t=1}^T\|\ell_t\|_*^2} \right)$
where the constant hidden in $O(\cdot)$ depends on the decision set and the
regularizer.  In Section~\ref{section:lower-bound}, we show that the
$\sqrt{\sum_{t=1}^T \|\ell_t\|_*^2}$ term is necessary by proving $\Omega
\left(D \sqrt{\sum_{t=1}^T\|\ell_t\|_*^2} \right)$ lower bound on the regret of
any algorithm for OLO for any decision set with diameter $D$ with respect to
the primal norm $\|\cdot\|$.

For the \textsc{SOLO FTRL} algorithm, we prove an $O \left(\max_{t=1,2,\dots,T}
\|\ell_t\|_* \sqrt{T} \right)$ regret bound for any unbounded decision set. The
constant hidden in $O(\cdot)$ depends not only on the decision set and the
regularizer, but also on the competitor $u \in K$.~\footnote{An upper bound
that does not depend on $u$ is impossible. It would contradict the $\Omega
\left( D \sqrt{\sum_{t=1}^T\|\ell_t\|_*^2} \right)$ lower bound.} This is the
\textbf{first adaptive algorithm for unbounded decision sets} with a
non-trivial upper bound.

@@

Our proofs of the regret upper bounds rely on homogeneous inequalities
(Lemmas~\ref{lemma:recurrence-solution},
\ref{lemma:sum-of-square-roots-inverses}, \ref{lemma:useful}).  Two of the
inequalities are new. Previously, many proofs of regret bounds relied on ad-hoc
non-homogeneous inequalities, or the existing algorithms used quantities (e.g.
$\sum_{t=1}^T \|\ell_t\|_*^2$) that are not available until all loss vectors
are processed.

Finally, in Section~\ref{subsection:mirror-descent-lower-bound}, we show
negative results for existing popular variants of \textsc{Mirror Descent}. We
show that variants of \textsc{Mirror Descent} have $\Omega(T)$ regret if the
Bregman divergence associated with the regularizer is unbounded on the decision
set.  The setting includes the standard gradient descent with step size
$1/\sqrt{t}$, \textsc{AdaGrad MD} and \textsc{Scale-Free Mirror Descent} on a
unbounded decision set. This result suggests that \textsc{Follow The
Regularized Leader} is superior to \textsc{Mirror Descent}.

>>>>

% \begin{definition}[Scale-Free Algorithm]
% \label{definition:scale-free-algorithm}
% An algorithm for OLO is \emph{scale-free} if for any sequence of loss vectors
% $\{\ell_t\}_{t=1}^\infty$ and any real number $c > 0$, on the sequences of
% $\{\ell_t\}_{t=1}^\infty$ and $\{c \cdot \ell_t\}_{t=1}^\infty$, the algorithm
% chooses the same sequence of points $\{w_t\}_{t=1}^\infty$.
% \end{definition}

The most important property of scale-free algorithms is that both their loss
and their regret is scales linearly with the maximum norm of the loss vector
appearing in the sequence.

\subsection{Previous results}

Majority of existing algorithms for OLO are based on two generic algorithms:
\textsc{Follow The Regularizer Leader} (\textsc{FTRL}) and \textsc{Mirror
Descent} (\textsc{MD}). \textsc{FTRL} dates back to the potential-based
forecaster in \cite[Chapters~11]{Cesa-Bianchi-Lugosi-2006} and its theory was
developed by \cite{Shalev-Shwartz-2007}.  The name \textsc{Follow The
Regularized Leader} comes from~\cite{Abernethy-Hazan-Rakhlin-2008}.
Independently, the same algorithm was proposed by~\cite{Nestorov-2009} for
convex optimization under the name \textsc{Dual Averaging} and later
rediscovered in~\cite{Xiao-2010} for online convex optimization. The general
version with time-varying regularizers was analyzed in
\cite{Orabona-Crammer-Cesa-Bianchi-2014}. \textsc{MD} was originally proposed
in \cite{Nemirovski-Yudin-1983} and later analyzed in~\cite{Beck-Teboulle-2003}
for convex optimization. In the online literature it makes the first
appearance, under a different name, in \cite{Kivinen-Warmuth-1997}. Good
description of \textsc{MD} is given in~\cite{Rakhlin-Sridharan-2009}.

Both \textsc{FTRL} and \textsc{MD} are parametrized by a function called a
\emph{regularizer}. The function is assumed to be strongly convex with respect
to a given norm $\|\cdot\|$. Based on different regularizers different
algorithms with different properties can be instantiated. A summary of
algorithms for OLO is presented in Table~\ref{table:results}.  All of them,
except for \textsc{Hedge}, are instances of \textsc{FTRL} or \textsc{MD}.

Scale-free versions of \textsc{MD} include \textsc{AdaGrad
MD}~\cite{Duchi-Hazan-Singer-2011} and~\textsc{Optimistic
MD}~\cite{Rakhlin-Sridharan-2013}. However, these algorithms have non-trivial
regret bounds only when the Bregman divergence associated with the regularizer
is bounded, which happens only when the decision is bounded as well. As we show
in Section~\ref{subsection:mirror-descent-lower-bound}, this is unavoidable,
since \textsc{AdaGrad MD} and similar algorithms based on \textsc{MD} can have
$\Omega(T)$ regret if Bregman divergence is not bounded.

Only one scale-free algorithm based on \textsc{FTRL} was known. It is the
\textsc{AdaHedge}~\cite{de-Rooij-van-Erven-Grunwald-Koolen-2014} algorithm for
learning with expert advice, where the decision set is bounded.
Algorithms based on \textsc{FTRL} that are ``almost'' scale-free include
\textsc{FTRL-Proximal}~\cite{McMahan-Streeter-2010,McMahan-2014} and
\textsc{AdaGrad FTRL}~\cite{Duchi-Hazan-Singer-2011}.  These algorithms fail to
be scale-free due to ``off-by-one'' issue; see~\cite{McMahan-2014} and also our
discussion in Section~\ref{section:solo-ftrl}.

For unbounded decision sets no scale-free algorithm with a non-trivial regret
bound was known. Unbounded decision sets are practically important, since
learning large-scale linear models (e.g. logistic regression) can be reduced to
OLO with $\R^d$ as the decision set. One of our contributions is an
\textsc{FTRL}-based scale-free algorithm with a non-trivial regret bound for
any regularizer and any decision set, bounded or unbounded.

% While the scale-free property has been looked at in the expert setting, in
% the general OLO setting this issue has been largely ignored.  In particular,
% the \textsc{AdaHedge}~\cite{de-Rooij-van-Erven-Grunwald-Koolen-2014}
% algorithm, for prediction with expert advice, is specifically designed to be
% scale-free.  A notable exception in the OLO literature is the discussion of
% the ``off-by-one'' issue in~\cite{McMahan-2014}, where it is explained that
% even the popular \textsc{AdaGrad} algorithm~\cite{Duchi-Hazan-Singer-2011} is
% not completely adaptive; see also our discussion in
% Section~\ref{section:solo-ftrl}. In particular, existing scale-free
% algorithms cover only some norms/regularizers and \emph{only} bounded
% decision sets. The case of \textbf{unbounded decision sets}, practically the
% most interesting one for machine learning applications, remains completely
% unsolved.

\subsection{Overview of the Results}

We design and analyze three scale-free algorithms: \textsc{AdaFTRL},
\textsc{SOLO FTRL} and \textsc{Scale-Free MD}.  \textsc{AdaFTRL} and
\textsc{SOLO FTRL} are based on \textsc{FTRL}.  \textsc{AdaFTRL} is a
generalization of
\textsc{AdaHedge}~\cite{de-Rooij-van-Erven-Grunwald-Koolen-2014} to arbitrary
strongly convex regularizers.  \textsc{SOLO FTRL} can be viewed as the
``correct'' scale-free version of \textsc{AdaGrad
FTRL}~\cite{Duchi-Hazan-Singer-2011} generalized to arbitrary strongly convex
regularizers.  \textsc{Scale-Free MD} is based on \textsc{MD}. It is a
generalization of \textsc{AdaGrad MD}~\cite{Duchi-Hazan-Singer-2011} to
arbitrary strongly convex regularizers, similar to \textsc{Optimistic
MD}~\cite{Rakhlin-Sridharan-2013}.  The three algorithms are presented and
analyzed in Sections~\ref{section:ada-ftrl}, \ref{section:solo-ftrl} and
\ref{section:mirror-descent} respectively.

We prove that the regret of \textsc{AdaFTRL} and \textsc{SOLO FTRL} after $T$
rounds is bounded by $O \left(\sqrt{\sup_{v \in K} f(v)
\sum_{t=1}^T\|\ell_t\|_*^2} \right)$ where $f$ is a non-negative regularizer that is
$1$-strongly convex with respect to a norm $\|\cdot\|$ and $\|\cdot\|_*$ is its
dual norm. For \textsc{Scale-Free MD}, we prove $O \left(\sqrt{\sup_{u,v \in K}
B_f(u,v) \sum_{t=1}^T\|\ell_t\|_*^2} \right)$ where $B_f$ is the Bregman
divergence associated with a $1$-strongly convex regularizer $f$. In
Section~\ref{section:lower-bound}, we show that the $\sqrt{\sum_{t=1}^T
\|\ell_t\|_*^2}$ term in the bounds term is necessary by proving
$\frac{D}{\sqrt{8}} \sqrt{\sum_{t=1}^T\|\ell_t\|_*^2}$ lower bound on the
regret of any algorithm for OLO for any decision set with diameter $D$ with
respect to the primal norm $\|\cdot\|$.

For \textsc{SOLO FTRL}, we prove that regret against a competitor $u \in K$ is
at most $O \left(f(u) \sqrt{\sum_{t=1}^T \|\ell_t\|_*^2} + \max_{t=1,2,\dots,T}
\|\ell_t\|_* \sqrt{T} \right)$.  Same as before, $f$ is a non-negative
$1$-strongly convex regularizer. The decision can be bounded or unbounded. This
is the \textbf{first adaptive algorithm for unbounded decision sets} with a
non-trivial upper bound.

@@

Our proofs of the regret upper bounds rely on homogeneous inequalities
(Lemmas~\ref{lemma:recurrence-solution}, \ref{lemma:useful},
\ref{lemma:sum-of-square-roots-inverses}).  Previously, many proofs of regret
bounds relied on ad-hoc non-homogeneous inequalities, or the existing
algorithms used quantities (e.g.  $\sum_{t=1}^T \|\ell_t\|_*^2$) that are not
available until all loss vectors are processed. Two of the inequalities
(Lemmas~\ref{lemma:recurrence-solution} and \ref{lemma:useful}) are new and
interesting in their own right.

Finally, in Section~\ref{subsection:mirror-descent-lower-bound}, we show
negative results for existing popular variants of \textsc{MD}. We
show that variants of \textsc{MD} have $\Omega(T)$ regret if the
Bregman divergence associated with the regularizer is unbounded.
The setting includes the standard gradient descent with step size
$1/\sqrt{t}$, \textsc{AdaGrad MD} and \textsc{Scale-Free MD} on a
unbounded decision set. This result suggests that \textsc{FTRL}
is superior to \textsc{MD}.

### 2015-12-11 mirror-descent.tex
Expand introduction

\textsc{Mirror Descent}\footnote{The algorithm was originally proposed in
\cite{Nemirovski-Yudin-1983} and later analyzed in~\cite{Beck-Teboulle-2003} for convex optimization. In the online literature it makes the first appeareance, under a different name, in \cite{Kivinen-Warmuth-1997}.}
is a generic algorithm similar to \textsc{Follow The Regularized Leader} but
quite different in details. The algorithm is stated as
Algorithm~\ref{algorithm:mirror-descent-varying-regularizer}. The algorithm is
parametrized by a sequence $\{R_t\}_{t=0}^\infty$ of convex functions $R_t:K
\to \R$ called \emph{regularizers}. Each regularizer $R_t$ can depend on past
loss vectors $\ell_1, \ell_2, \dots, \ell_t$ in an arbitrary way. If $R_t$ is
not differentiable\footnote{Note that this can happen even when $R_t$ is a
restriction of a differentiable function defined on a superset of $K$.  If $K$
is bounded and closed, $R_t$ fails to be differentiable at the boundary of $K$.
If $K$ is a subset of an affine subspace of a dimension smaller than the
dimension of $V$, then $R_t$ fails to be differentiable everywhere.}, the Bregman
divergence, $\Breg_{R_t}(u,v) = R(u) - R(v) - \langle \grad R(v), u - v
\rangle$ needs to be defined. This is done by choosing a subgradient map $\grad
R_t:K \to V$, i.e. a function such that $\grad R_t(w)$ is a subgradient of
$R_t$ at a point $w$. If $R_t$ is a restriction of a differentiable function
$R'_t$, it is convenient to define $\grad R_t(w) = \grad R'_t(w)$ for all $w
\in K$. The following lemma bounds regret of \textsc{Mirror Descent}.  Proof
can be found
in~\cite{Rakhlin-Sridharan-2009,Duchi-Shalev-Shwartz-Singer-Tewari-2010}.  For
completeness we give a proof in~\ref{section:mirror-descent-proofs}.

>>>>

\textsc{Mirror Descent} is a generic algorithm similar to \textsc{Follow The Regularized Leader} but
quite different in details. The algorithm is stated as
Algorithm~\ref{algorithm:mirror-descent-varying-regularizer}. The algorithm is
parametrized by a sequence $\{R_t\}_{t=0}^\infty$ of convex functions $R_t:K
\to \R$ called \emph{regularizers}. Each regularizer $R_t$ can depend on past
loss vectors $\ell_1, \ell_2, \dots, \ell_t$ in an arbitrary way. If $R_t$ is
not differentiable\footnote{Note that this can happen even when $R_t$ is a
restriction of a differentiable function defined on a superset of $K$.  If $K$
is bounded and closed, $R_t$ fails to be differentiable at the boundary of $K$.
If $K$ is a subset of an affine subspace of a dimension smaller than the
dimension of $V$, then $R_t$ fails to be differentiable everywhere.}, the Bregman
divergence, $\Breg_{R_t}(u,v) = R(u) - R(v) - \langle \grad R(v), u - v
\rangle$ needs to be defined. This is done by choosing a subgradient map $\grad
R_t:K \to V$, i.e. a function such that $\grad R_t(w)$ is a subgradient of
$R_t$ at a point $w$. If $R_t$ is a restriction of a differentiable function
$R'_t$, it is convenient to define $\grad R_t(w) = \grad R'_t(w)$ for all $w
\in K$. The following lemma bounds regret of \textsc{Mirror Descent}.  Proof
can be found
in~\cite{Rakhlin-Sridharan-2009,Duchi-Shalev-Shwartz-Singer-Tewari-2010}.  For
completeness we give a proof in~\ref{section:mirror-descent-proofs}.

### 2015-12-11 preliminaries.tex
Expand introduction

Our scale-free online learning algorithms are versions of the \textsc{Follow
The Regularized Leader}\footnote{The name \textsc{Follow The Regularized
Leader} comes from~\cite{Abernethy-Hazan-Rakhlin-2008}. It dates back to the potential-based forecaster in \cite[Chapters~11]{Cesa-Bianchi-Lugosi-2006} and its theory was heavily developed by \cite{Shalev-Shwartz-2007}. 
Independently, the same algorithm was proposed by~\cite{Nestorov-2009} for convex optimization under the name \textsc{Dual
Averaging} and later rediscovered in~\cite{Xiao-2010} for online convex optimization. The version with time-varying regularizers was analyzed in \cite{Orabona-Crammer-Cesa-Bianchi-2014}.
} (\textsc{FTRL})
algorithm with \emph{varying regularizers}, presented as
Algorithm~\ref{algorithm:ftrl-varying-regularizer}. The algorithm is
paramatrized by a sequence $\{R_t\}_{t=1}^\infty$ of convex functions $R_t:K
\to \R$ called \emph{regularizers}.  Each regularizer $R_t$ can depend on the
past loss vectors $\ell_1, \ell_2, \dots, \ell_{t-1}$ in an arbitrary way.  The
following lemma bounds its regret.

>>>>

Our scale-free online learning algorithms are versions of the \textsc{Follow
The Regularized Leader} (\textsc{FTRL}) algorithm with \emph{varying
regularizers}, presented as Algorithm~\ref{algorithm:ftrl-varying-regularizer}.
The algorithm is paramatrized by a sequence $\{R_t\}_{t=1}^\infty$ of convex
functions $R_t:K \to \R$ called \emph{regularizers}.  Each regularizer $R_t$
can depend on the past loss vectors $\ell_1, \ell_2, \dots, \ell_{t-1}$ in an
arbitrary way.  The following lemma bounds its regret.
